Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-09 analysis/paper/paper.Rmd
added code to replicate prediktera PCA figures and loadings with ggplot/ggbiplot

```{r load-packages-and-data}
library(tidyverse)
library(gridExtra)
library(ggplot2)
library(gridExtra)
library(ggsci)
library(prospectr)
library(patchwork)
library(factoextra)
library(ggbiplot)

#Import xrf data
XRF.csv <-
  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE)

@@

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
Points.xrf[8][is.na(Points.xrf[8])] <- "Translucent"

#prepare color/fill and symbols for score plots
colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
shapes <- c(24,25,23,22,21)

@@

#prepare labels for PCs
pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = pc1lab,
       y = pc2lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```
 \linebreak

@@

#prepare labels for PCs
pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 1 000 - 2 500 nm") +
  labs(x = pc1lab,
       y = pc2lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

```{r raman-spectra, fig.width=20, fig.height=8, fig.cap="Spectra of raman measurements sampled on a dark quartzite featuring both the *quartz peak* at 460 cm-1, as well as the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band)"}
#Identify file with raman spectra and import
file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
spectra <- read.csv2(file, sep = ";", dec = ",", header = TRUE)

@@

plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
           xlab = "Wavenumber (cm-1)", ylab = "Intensity")
```

@@

#prepare labels for PCs
pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
pc1lab <- paste0("PC1 (",as.character(pc1var),"%)")
pc2lab <- paste0("PC2 (",as.character(pc2var),"%)")
pc3lab <- paste0("PC3 (",as.character(pc3var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Raman") +
  labs(x = pc1lab,
       y = pc2lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

#bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  labs(x = pc1lab,
       y = pc3lab) +
  scale_shape_manual(name = "Material",
                     values=shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

comb <-
  p1 + p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                  axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                  axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                  legend.title = element_text(size = 12, face = "bold", colour = "black"),
                  legend.background = element_rect(linetype = "solid", color = "black"),
                  legend.position = "bottom")

@@

#Prepare colors
colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")

@@

#Violin plot of elemental content, trimmed, with fill based on elemental groups
Points_long %>%
  drop_na(Value) %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=colors) +
    theme_bw() +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#bar graph of the loadings (P1-P3) for the XRF data  
loadings %>%
  ggplot(aes(x=fct_rev(Element), y=Value)) +
    geom_bar(
      stat = "identity", position = "identity",
      color = "black", fill = "darkgrey",
      width = 0.5) +
  facet_wrap( ~ P, scales = "free") +
  theme_bw() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10, face="bold", colour = "black"),
        axis.title.x = element_blank(),
        strip.text.x = element_text(size=14, face="bold", colour="black")) +
  coord_flip()

```

@@

<!--```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
knitr::include_graphics('./analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
```-->

>>>>

```{r load-packages-and-data}
library(ChemoSpec)
library(data.table)
library(factoextra)
library(ggbiplot)
library(ggplot2)
library(ggsci)
library(gridExtra)
library(missMDA)
library(patchwork)
library(prospectr)
library(tidyverse)

#Import xrf data
XRF.csv <-
  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, na = c("","NA","NULL"))

@@

#Fill the NA fields in the munsell_hue column and mark them as translucent samples
Points.nir[8][is.na(Points.nir[8])] <- "Translucent"
Points.raman[8][is.na(Points.raman[8])] <- "Translucent"
Points.xrf[20][is.na(Points.xrf[20])] <- "Translucent"

#prepare color/fill and symbols for score plots
pca.colors <- c("#E31CFF", "#718EFF", "#0080FF", "#D5DF9E", "#C6DF39", "#FFFFC1", "#FFFFA1", "#FFFF00", "#FFFF38", "#FFB468", "#000000", "#202020", "#404040", "#606060", "#808080", "#9F9F9F", "#DFDFDF", "#FFFFFF")
pca.shapes <- c(24,25,23,22,21)
pca.munsell_hue <- c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9")

@@

#prepare labels for PCs
vis.pc1var <- round(summary(visnir.pca)$importance[2,1]*100, digits=1)
vis.pc2var <- round(summary(visnir.pca)$importance[2,2]*100, digits=1)
vis.pc1lab <- paste0("PC1 (",as.character(vis.pc1var),"%)")
vis.pc2lab <- paste0("PC2 (",as.character(vis.pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 400 - 750 nm") +
  labs(x = vis.pc1lab,
       y = vis.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", 
                             title.hjust = 0.5)) +
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```
 \linebreak

@@

#prepare labels for PCs
nir.pc1var <- round(summary(nir.pca)$importance[2,1]*100, digits=1)
nir.pc2var <- round(summary(nir.pca)$importance[2,2]*100, digits=1)
nir.pc1lab <- paste0("PC1 (",as.character(nir.pc1var),"%)")
nir.pc2lab <- paste0("PC2 (",as.character(nir.pc2var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
ggplot(cbind(basic_plot1$data, Points.nir[, c(8,10)]),
             aes(x=x, y=y, 
                 shape = material, 
                 fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Near infrared 1 000 - 2 500 nm") +
  labs(x = nir.pc1lab,
       y = nir.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(plot.title = element_text(size = 12, face = "bold", colour = "black", vjust = - 10, hjust = 0.02),
        axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
        axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
        legend.title = element_text(size = 12, face = "bold", colour = "black"),
        legend.background = element_rect(linetype = "solid", color = "black"),
        legend.position = "bottom")
```

@@

```{r raman-spectra, fig.width=20, fig.height=10, fullwidth = TRUE, fig.cap="Mean spectrum of the raman data grouped by their hue. The *quartz peak* at 460 cm-1 is present in all four groups, but the *graphite peaks* at 1 287  cm-1 (D band) and 1 598 cm-1 (G band) mainly appear in the light and dark groups"}
#Read and create new file for chemospec
raman <-matrix2SpectraObject(
  gr.crit = c("Translucent", "White", "Dark", "Light" ),
  gr.cols = c("purple", "blue", "black", "yellow"),
  freq.unit ="Wavenumber (cm-1)",
  int.unit ="Intensity",
  descrip ="Bifacial points measurements",
  in.file = "./analysis/data/derived_data/raman_baseline_corrected_transposed_20210908.csv",
  sep = ";",
  dec = ",",
  chk = TRUE,
  out.file = "raman")

surveySpectra(raman, by.gr = TRUE , method = "sem", main = "mean")

```
<!-- #Identify file with raman spectra and import
file <- '../quantifyingquartzite/analysis/data/derived_data/raman_baseline_corrected_transposed_20210826.csv'
spectra <- read.csv2(file, sep = ";", dec = ",", header = TRUE)

@@

plot_multi(spectra, y=120, cols = rainbow(ncol(spectra)), type='l', lwd=3,
           xlab = "Wavenumber (cm-1)", ylab = "Intensity") -->

@@

#prepare labels for PCs
raman.pc1var <- round(summary(raman.pca)$importance[2,1]*100, digits=1)
raman.pc2var <- round(summary(raman.pca)$importance[2,2]*100, digits=1)
raman.pc3var <- round(summary(raman.pca)$importance[2,3]*100, digits=1)
raman.pc1lab <- paste0("PC1 (",as.character(raman.pc1var),"%)")
raman.pc2lab <- paste0("PC2 (",as.character(raman.pc2var),"%)")
raman.pc3lab <- paste0("PC3 (",as.character(raman.pc3var),"%)")

#bind the basic fviz plot for PC 1 and 2, and use as basis for a more customizeable plot in ggpplot
raman.p1 <- ggplot(cbind(basic_plot1$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  ggtitle("Raman") +
  labs(x = raman.pc1lab,
       y = raman.pc2lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

#bind the basic fviz plot for PC 1 and 3, and use as basis for a more customizeable plot in ggpplot
raman.p2 <- ggplot(cbind(basic_plot2$data, Points.raman[, c(8,10)]),
       aes(x=x, y=y, 
           shape = material, 
           fill = munsell_hue)) +
  geom_point(size=3) +
  theme_bw() +
  labs(x = raman.pc1lab,
       y = raman.pc3lab) +
  scale_shape_manual(name = "Material",
                     values=pca.shapes) +
  scale_fill_manual(name = "Munsell hue",
                    breaks = c("Translucent", "2.5PB", "10B", "2.5GY", "5GY", "2.5Y", "5Y", "7.5Y", "10Y", "5YR", "N1", "N2", "N3", "N4", "N5", "N6", "N8", "N9"),
                    values=pca.colors) +
  guides(shape = guide_legend(override.aes = list(fill = "black"), 
                              ncol=2, 
                              title.position="top", 
                              title.hjust = 0.5),
         fill = guide_legend(override.aes = list(shape = 22,
                                                 fill = pca.colors,
                                                 color = "black",
                                                 size=3),
                             title.position="top", title.hjust = 0.5))+
  theme(legend.position = "none")

comb <-
  raman.p1 + raman.p2 & theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
                              axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
                              legend.title = element_text(size = 12, face = "bold", colour = "black"),
                              legend.background = element_rect(linetype = "solid", color = "black"),
                              legend.position = "bottom")

@@

#Prepare colors
loading.colors <- c("darkorange", "darkolivegreen3", "skyblue1", "lightslateblue", "salmon1", "lightgoldenrod1")

@@

#Violin plot of elemental content, trimmed, with fill based on elemental groups
Points_long %>%
  drop_na(Value) %>%
    ggplot(aes(x=Element, y=Value, fill = Group)) +
    geom_violin(trim = TRUE) +
    facet_wrap( ~ Element, scales = "free", labeller = labeller(Element = Elements)) +
    scale_fill_manual(values=loading.colors) +
    theme_bw() +
    theme(
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      strip.text.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_blank(),
      plot.title = element_text(size=10, face = "bold", colour = "black", margin = margin(t = 10, b = -20), vjust=0.02, hjust=0.01),
      axis.title.y=element_blank(),
      axis.text.x=element_blank(),
      axis.ticks.x=element_blank())
```

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#select elements to include in pca, as well as columns to group by
xrf <-
  Points.xrf %>%
    select(reading_no, sample_id, munsell_hue, material, X12Mg, X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti)

#log transform data
xrf.log <- log(xrf[,5:20])

#impute missing values
xrf.imp <- imputePCA(xrf.log, ncp = 2, scale = FALSE, method = "Regularized", maxiter = 1000)

#perform pca with mean centering and unite variance scaling
xrf.pca <- prcomp(xrf.imp$completeObs, center = TRUE, scale.=TRUE)

#Extract loadings from xrf.pca
loadings <- as.data.frame(xrf.pca$rotation[,1:3])
#As elements are stored as rownames, set them as first column so it can be used as a variable
loadings <- setDT(loadings, keep.rownames = "Element")[]
# remove inital 'X' from element names, which was introduced in initial import to R
loadings$Element <- substring(loadings$Element, 2)
#Melt loadings from wide to long format
loadings.melted <- melt(loadings, id.vars="Element")
#Set names of columns
loadings <- setNames(loadings.melted, c("Element", "PC", "Value"))

#bar graph of the loadings (P1-P3) for the XRF data
loadings %>%
  ggplot(aes(x=fct_rev(Element), y=Value)) +
  geom_bar(
    stat = "identity", position = "identity",
    color = "black", fill = "darkgrey",
    width = 0.5) +
  facet_wrap( ~ PC, scales = "free") +
  theme_bw() +
  theme(plot.margin = margin(10,10,10,10),
        axis.title.y = element_blank(),
        axis.text.y = element_text(size=10, face="bold", colour = "black"),
        axis.title.x = element_blank(),
        strip.text.x = element_text(size=14, face="bold", colour="black")) +
  coord_flip()
```

@@

<!--```{r xrf-score-plot-p1-p2, fig.cap="Score plot for P1 and P2, the two major groups are distinguished by lower SiO content to the left and higher to the right"}
knitr::include_graphics('./analysis/figures/fig_4_7_xrf_pca_p1_p1.jpg')
```-->
```{r xrf-pca1, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 2nd components. The smaller group to the left in the plot includes samples with higher Si content, while the larger group includes samples with Si content < 50%."}

#Prepare two score plots
ggbiplot(xrf.pca,
         choices=c(1,2),
         obs.scale = 1,
         var.scale = 1,
         group=NULL,
         varname.size = 4,
         labels.size=3,
         ellipse = FALSE,
         circle = FALSE) +
theme_minimal() +
geom_point(aes(shape=Points.xrf$material, 
               fill = Points.xrf$munsell_hue), 
           size = 3) +
coord_cartesian() +
scale_shape_manual(name = "Material",
                   values=pca.shapes) +
scale_fill_manual(name = "Munsell hue",
                  breaks = pca.munsell_hue,
                  values=pca.colors) +
guides(shape = guide_legend(override.aes = list(fill = "black"),
                            ncol=2,
                            title.position="top",
                            title.hjust = 0.5),
       fill = guide_legend(override.aes = list(shape = 22,
                                               fill = pca.colors,
                                               color = "black",
                                               size=3),
                           title.position="top", title.hjust = 0.5)) +
theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      legend.background = element_rect(linetype = "solid", color = "black"),
      legend.position = "bottom")
```

```{r xrf-pca2, fig.width=15, fullwidth=TRUE, fig.height=10, fig.cap="XRF score plot of 1^st^ and 3rd components. The same groupings of low/high Si content remains along PC1, but PC3 seems to separate darker and light samples further in the low Si group."}
ggbiplot(xrf.pca,
         choices=c(1,3),
         obs.scale = 1,
         var.scale = 1,
         group=NULL,
         varname.size = 4,
         labels.size=3,
         ellipse = FALSE,
         circle = FALSE) +
theme_minimal() +
geom_point(aes(shape=Points.xrf$material, 
               fill = Points.xrf$munsell_hue), 
           size = 3) +
coord_cartesian() +
scale_shape_manual(name = "Material",
                   values=pca.shapes) +
scale_fill_manual(name = "Munsell hue",
                  breaks = pca.munsell_hue,
                  values=pca.colors) +
guides(shape = guide_legend(override.aes = list(fill = "black"),
                            ncol=2,
                            title.position="top",
                            title.hjust = 0.5),
       fill = guide_legend(override.aes = list(shape = 22,
                                               fill = pca.colors,
                                               color = "black",
                                               title.position="top", 
                                               title.hjust = 0.5))) +
theme(plot.title = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.x = element_text(size = 12, face = "bold", colour = "black"),
      axis.title.y = element_text(size = 12, face = "bold", colour = "black"),
      legend.title = element_text(size = 12, face = "bold", colour = "black"),
      legend.background = element_rect(linetype = "solid", color = "black"),
      legend.position = "bottom")
```
